# Attempt 2 — Richness Slope Modelling: Preprocessing
For each BDM plot, compute OLS richness slopes (species/year) and per-month
climatic trends (2001–2025) as predictors. One row per plot:
`total_change_pct` ~ climate slopes + baselines + static env

# Load Libraries
```{r}
library(tidyverse)
library(sf)
library(DataExplorer)
library(ggcorrplot)

source("_config.r")
```

# Richness Data
```{r}
moss_richness  <- readRDS(PATH_MOSS_RICHNESS)   # aID_STAO, yearMoos, richness_moss
plant_richness <- readRDS(PATH_PLANT_RICHNESS)  # aID_STAO, yearPl,   richness_plant

head(moss_richness)
head(plant_richness)

dim(moss_richness)
dim(plant_richness)
```

# Climate Panel
```{r}
clim_panel <- readRDS(PATH_CLIM_MONTHLY_PANEL)  # aID_STAO, year, month, tabs, tmax, tmin, rhires, srel, sisdir

head(clim_panel)
cat("Climate panel dimensions:", dim(clim_panel), "\n")
cat("Years available:", min(clim_panel$year), "–", max(clim_panel$year), "\n")
```

# Static Environmental Data
## Load Environmental Data
```{r}
# Climate comes from the panel (slopes + baseline), not from the aggregated MeteoSwiss rasters
env_arealstatistik   <- readRDS(PATH_ENV_LANDUSE)
env_dhm25            <- readRDS(PATH_ENV_TOPO)
env_habitatmap       <- readRDS(PATH_ENV_HABITAT)
env_hab_landscape    <- readRDS(PATH_ENV_HAB_WIDE_2M)
env_soil_kobo_interp <- readRDS(PATH_ENV_SOIL_KOBO_INTERP)

head(env_arealstatistik)
head(env_dhm25)
head(env_habitatmap)
head(env_hab_landscape)
head(env_soil_kobo_interp)

plot_coordinates <- readRDS(PATH_PLOT_COORDS_RDS)
cat("Number of rows in plot_coordinates:", nrow(plot_coordinates), "\n")
cat("Number of rows in env_arealstatistik:", nrow(env_arealstatistik), "\n")
cat("Number of rows in env_dhm25:", nrow(env_dhm25), "\n")
cat("Number of rows in env_habitatmap:", nrow(env_habitatmap), "\n")
cat("Number of rows in env_hab_landscape:", nrow(env_hab_landscape), "\n")
cat("Number of rows in env_soil_kobo_interp:", nrow(env_soil_kobo_interp), "\n")
```

## Join Environmental Data
```{r}
env_data <- plot_coordinates |>
  left_join(env_arealstatistik,   by = "aID_STAO") |>
  left_join(env_dhm25,            by = "aID_STAO") |>
  left_join(env_habitatmap,       by = "aID_STAO") |>
  left_join(env_hab_landscape,    by = "aID_STAO") |>
  left_join(env_soil_kobo_interp, by = "aID_STAO")
View(env_data)
```

## Turn numerics into binned categories
```{r}
env_data <- env_data |>
  mutate(aspect_factor = factor(
    case_when(
      aspect >= 337.5 | aspect < 22.5  ~ "N",
      aspect >= 22.5  & aspect < 67.5  ~ "NE",
      aspect >= 67.5  & aspect < 112.5 ~ "E",
      aspect >= 112.5 & aspect < 157.5 ~ "SE",
      aspect >= 157.5 & aspect < 202.5 ~ "S",
      aspect >= 202.5 & aspect < 247.5 ~ "SW",
      aspect >= 247.5 & aspect < 292.5 ~ "W",
      aspect >= 292.5 & aspect < 337.5 ~ "NW"
    ),
    levels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW")))

env_data$aspect_factor
```

## Turn categorical variables into factors
```{r}
env_data <- env_data |>
  mutate(across(where(is.character), as.factor))

typoch_columns <- grep("TypoCH", names(env_data), value = TRUE)

env_data <- env_data |>
  mutate(across(all_of(c("AS_4", "LU_4", "LU_10", typoch_columns)), as.factor))

column_types <- sapply(env_data, class)
print(column_types)
```

## Reduce the number of columns
```{r}
env_data <- env_data |>
  select(-XKoord_LV95, -YKoord_LV95) |>
  select(-matches("TypoCH_NUM"), -matches("TypoCH_DE_1st"), -matches("TypoCH_DE_2nd")) |>
  select(-matches("TypoCH_2nd_code")) |>
  select(-LU_4, -AS_4) |>
  select(-slope)
```

# Compute per-plot climate features (slopes + baseline)
```{r}
compute_climate_features <- function() {
  clim_vars <- c("tabs", "tmax", "tmin", "rhires", "srel", "sisdir")

  clim_panel |>
    filter(year >= 2001 & year <= 2025) |>
    tidyr::pivot_longer(cols = all_of(clim_vars), names_to = "clim_var", values_to = "value") |>
    group_by(aID_STAO, clim_var, month) |>
    summarise(
      slope    = coef(lm(value ~ year))[["year"]],
      baseline = mean(value, na.rm = TRUE),
      .groups  = "drop"
    ) |>
    tidyr::pivot_wider(
      names_from  = c(clim_var, month),
      values_from = c(slope, baseline),
      names_glue  = "{.value}_{clim_var}_m{month}"
    )
}

clim_features <- compute_climate_features()

cat("Climate features:", nrow(clim_features), "×", ncol(clim_features), "\n")
```

# Compute richness slopes and target variable
```{r}
moss_richness_summary <- moss_richness |>
  group_by(aID_STAO) |>
  filter(n() >= 3) |>
  summarise(
    slope_year        = coef(lm(richness_moss ~ yearMoos))[["yearMoos"]],
    intercept         = coef(lm(richness_moss ~ yearMoos))[["(Intercept)"]],
    n_obs             = n(),
    year_min          = min(yearMoos, na.rm = TRUE),
    year_max          = max(yearMoos, na.rm = TRUE),
    mean_richness     = mean(richness_moss, na.rm = TRUE),
    baseline_richness = richness_moss[which.min(yearMoos)],
    .groups           = "drop"
  ) |>
  mutate(
    fitted_start     = intercept + slope_year * year_min,
    fitted_end       = intercept + slope_year * year_max,
    total_change_pct = ifelse(mean_richness == 0, 0,
                              ((fitted_end - fitted_start) / mean_richness) * 100)
  )

plant_richness_summary <- plant_richness |>
  group_by(aID_STAO) |>
  filter(n() >= 3) |>
  summarise(
    slope_year        = coef(lm(richness_plant ~ yearPl))[["yearPl"]],
    intercept         = coef(lm(richness_plant ~ yearPl))[["(Intercept)"]],
    n_obs             = n(),
    year_min          = min(yearPl, na.rm = TRUE),
    year_max          = max(yearPl, na.rm = TRUE),
    mean_richness     = mean(richness_plant, na.rm = TRUE),
    baseline_richness = richness_plant[which.min(yearPl)],
    .groups           = "drop"
  ) |>
  mutate(
    fitted_start     = intercept + slope_year * year_min,
    fitted_end       = intercept + slope_year * year_max,
    total_change_pct = ifelse(mean_richness == 0, 0,
                              ((fitted_end - fitted_start) / mean_richness) * 100)
  )

cat("Moss  richness summary:", nrow(moss_richness_summary),  "plots\n")
cat("Plant richness summary:", nrow(plant_richness_summary), "plots\n")
cat("Moss  NA in total_change_pct:", sum(is.na(moss_richness_summary$total_change_pct)), "\n")
cat("Plant NA in total_change_pct:", sum(is.na(plant_richness_summary$total_change_pct)), "\n")
```

# Assemble richness slopes datasets
```{r}
moss_richness_slopes_data <- moss_richness_summary |>
  select(aID_STAO, total_change_pct, baseline_richness) |>
  left_join(clim_features, by = "aID_STAO") |>
  left_join(env_data,      by = "aID_STAO") |>
  select(-aID_STAO)

plant_richness_slopes_data <- plant_richness_summary |>
  select(aID_STAO, total_change_pct, baseline_richness) |>
  left_join(clim_features, by = "aID_STAO") |>
  left_join(env_data,      by = "aID_STAO") |>
  select(-aID_STAO)

cat("Moss  richness slopes dataset:", nrow(moss_richness_slopes_data),  "×", ncol(moss_richness_slopes_data),  "\n")
cat("Plant richness slopes dataset:", nrow(plant_richness_slopes_data), "×", ncol(plant_richness_slopes_data), "\n")
cat("NA rows — moss: ",  sum(!complete.cases(moss_richness_slopes_data)),  "\n")
cat("NA rows — plant:", sum(!complete.cases(plant_richness_slopes_data)), "\n")

View(moss_richness_slopes_data)
```

# Save richness slopes datasets
```{r}
saveRDS(moss_richness_slopes_data,  PATH_MDL2_MOSS_DATA)
saveRDS(plant_richness_slopes_data, PATH_MDL2_PLANT_DATA)
```
