# Load Libraries
```{r}
library(tidyverse)
library(sf)
library(terra)
library(DataExplorer)
library(ggcorrplot)
```

# Richness Data processing
# Load Richness Data
```{r}
source("_config.r")

# Load alpha richness for mosses and plants
moss_richness <- readRDS(PATH_MOSS_RICHNESS)
plant_richness <- readRDS(PATH_PLANT_RICHNESS)

head(moss_richness)
head(plant_richness)

dim(moss_richness) # 7'105 x 3
dim(plant_richness) # 7'137 x 3
```

## Load Plot Coordinates
```{r}
# Load plot coordinates
plot_coordinates <- readRDS(PATH_PLOT_COORDS_RDS)

head(plot_coordinates)
```

# Environmental Data processing
## Load Environmental Data
```{r}
# Load extracted environmental variables
env_arealstatistik <- readRDS(PATH_ENV_LANDUSE)
env_dhm25 <- readRDS(PATH_ENV_TOPO)
env_habitatmap <- readRDS(PATH_ENV_HABITAT)
env_hab_landscape <- readRDS(PATH_ENV_HAB_WIDE_2M)
env_meteoswiss <- readRDS(PATH_ENV_CLIMATE)
env_soil_kobo_interp <- readRDS(PATH_ENV_SOIL_KOBO_INTERP)

head(env_arealstatistik)
head(env_dhm25)
head(env_habitatmap)
head(env_hab_landscape)
head(env_meteoswiss)
head(env_soil_kobo_interp)

# Print the amount of rows per environmental dataset to check if they match the plot coordinates
cat("Number of rows in plot_coordinates:", nrow(plot_coordinates), "\n")
cat("Number of rows in env_arealstatistik:", nrow(env_arealstatistik), "\n")
cat("Number of rows in env_dhm25:", nrow(env_dhm25), "\n")
cat("Number of rows in env_habitatmap:", nrow(env_habitatmap), "\n")
cat("Number of rows in env_hab_landscape:", nrow(env_hab_landscape), "\n")
cat("Number of rows in env_meteoswiss:", nrow(env_meteoswiss), "\n")
cat("Number of rows in env_soil_kobo_interp:", nrow(env_soil_kobo_interp), "\n")
```

## Join Environmental Data
```{r}
# Combine all static environmental baseline data and coordinates
env_data <- plot_coordinates |>
  left_join(env_arealstatistik, by = "aID_STAO") |>
  left_join(env_dhm25, by = "aID_STAO") |>
  left_join(env_habitatmap, by = "aID_STAO") |>
  left_join(env_hab_landscape, by = "aID_STAO") |>
  left_join(env_meteoswiss, by = "aID_STAO") |>
  left_join(env_soil_kobo_interp, by = "aID_STAO")
View(env_data)
```

## Turn numerics into binned categories
```{r}
# Bin aspect variable from dhm25 data
# Categories: North (340-50°), South-East (50-160°), South (160-250°), South-West (250-340°)
env_data <- env_data |>
  mutate(aspect_factor = factor(
    case_when(
      aspect >= 337.5 | aspect < 22.5  ~ "N",
      aspect >= 22.5  & aspect < 67.5  ~ "NE",
      aspect >= 67.5  & aspect < 112.5 ~ "E",
      aspect >= 112.5 & aspect < 157.5 ~ "SE",
      aspect >= 157.5 & aspect < 202.5 ~ "S",
      aspect >= 202.5 & aspect < 247.5 ~ "SW",
      aspect >= 247.5 & aspect < 292.5 ~ "W",
      aspect >= 292.5 & aspect < 337.5 ~ "NW"
    ), 
    levels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW")))

env_data$aspect_factor

# Done according to:
# https://www.supergeotek.com/SpatialAnalyst_ENG_HTML/aspect.htm
```

## Turn categorical variables into factors
```{r}
# Auto-convert all character columns to factors
env_data <- env_data |>
  mutate(across(where(is.character), as.factor))

# Grab all TypoCH columns
typoch_columns <- grep("TypoCH", names(env_data), value = TRUE)

# Convert other known categorical variables and typoch columns to factors
env_data <- env_data |>
  mutate(across(all_of(c("AS_4", "LU_4","LU_10", typoch_columns)), as.factor))

# Print the data type per column
column_types <- sapply(env_data, class)
print(column_types)
```

## Reduce the number of columns
Since i have a huge amount of predictors i try to reduce the number of columns by focusing on the 2nd level of TypoCH and Land usage.
Also i create data subsets based on the temporal aggregation levels in MeteoSwiss data.
I also remove columns which are not relevant for modelling such as coordinates, plotID etc.
```{r}
# Remove coordinates
env_data <- env_data |>
  select(-XKoord_LV95, -YKoord_LV95)

# Remove TypoCH text based columns
env_data <- env_data |>
  select(-matches("TypoCH_NUM"), -matches("TypoCH_DE_1st"), -matches("TypoCH_DE_2nd"))

# Remove TypoCH_2nd, only the first level is used in attempt 1
env_data <- env_data |>
  select(-matches("TypoCH_2nd_code"))

# Remove LU_4 and AS_4
env_data <- env_data |>
  select(-LU_4, -AS_4)

# Remove slope as this has been turned into factor
env_data <- env_data |>
  select(-slope)
```

## Combine with Richness Data
```{r}
# Join the environmental base data and filter to 2021-2025 directly
moss_model_data <- moss_richness |>
  filter(yearMoos >= 2021 & yearMoos <= 2025) |>
  left_join(env_data, by = "aID_STAO")
moss_model_data

plant_model_data <- plant_richness |>
  filter(yearPl >= 2021 & yearPl <= 2025) |>
  left_join(env_data, by = "aID_STAO")
plant_model_data

dim(moss_model_data) # 1'442 x 408
dim(plant_model_data) # 1'449 x 408
```

## Save the datasets
```{r}
# Save the combined datasets for modeling
saveRDS(moss_model_data, PATH_MDL1_MOSS_DATA)
saveRDS(plant_model_data, PATH_MDL1_PLANT_DATA)
```

# Explorative data analysis
## Clean categorical strings to prevent multibyte string errors in DataExplorer
```{r}
# Add the TypoCH_1st_DE column back to the dataset for EDA purposes
moss_model_data <- moss_model_data |>
  left_join(env_habitatmap |> select(aID_STAO, TypoCH_DE_1st), by = "aID_STAO")
plant_model_data <- plant_model_data |>
  left_join(env_habitatmap |> select(aID_STAO, TypoCH_DE_1st), by = "aID_STAO")

# Transform the column to factor
moss_model_data <- moss_model_data |>
  mutate(TypoCH_DE_1st = as.factor(TypoCH_DE_1st))
plant_model_data <- plant_model_data |>
  mutate(TypoCH_DE_1st = as.factor(TypoCH_DE_1st))

# Clean all categorical strings to strictly use ASCII characters
# First transliterate common Swiss accents/umlauts, then replace any remaining special characters with _
# This prevents invalid multibyte string errors in make.names() during DataExplorer report generation
umlaut_map <- c("ä" = "ae", "ö" = "oe", "ü" = "ue",
                "Ä" = "Ae", "Ö" = "Oe", "Ü" = "Ue",
                "é" = "e",  "è" = "e",  "ê" = "e",
                "à" = "a",  "â" = "a",  "ç" = "c")
clean_str <- function(x) stringr::str_replace_all(stringr::str_replace_all(x, umlaut_map), "[^a-zA-Z0-9_]", "_")

moss_model_data <- moss_model_data |>
  mutate(across(where(is.character), clean_str)) |>
  mutate(across(where(is.factor), ~ {
    factor(.x, levels = levels(.x), labels = clean_str(levels(.x)))
  }))
  s
plant_model_data <- plant_model_data |>
  mutate(across(where(is.character), clean_str)) |>
  mutate(across(where(is.factor), ~ {
    factor(.x, levels = levels(.x), labels = clean_str(levels(.x)))
  }))
```

## Select only yearly aggregated variables for the year 2021-2025
```{r}
# Reduce the dataset by selecting only relevant features for EDA
moss_model_data_reduced <- moss_model_data |>
  select(-aID_STAO, -matches("_m\\d{2}_"), -ends_with("_250"), -ends_with("_50"), -ends_with("_25"), -matches("_20yr"), -matches("_5yr"), -matches("TypoCH_NUM"), -matches("TypoCH_1st_code"), -matches("TypoCH_2nd_code"), -aspect)
View(moss_model_data_reduced)
dim(moss_model_data_reduced) # 1442 x 50

plant_model_data_reduced <- plant_model_data |>
  select(-aID_STAO, -matches("_m\\d{2}_"), -ends_with("_250"), -ends_with("_50"), -ends_with("_25"), -matches("_20yr"), -matches("_5yr"), -matches("TypoCH_NUM"), -matches("TypoCH_1st_code"), -matches("TypoCH_2nd_code"), -aspect)
View(plant_model_data_reduced)
dim(plant_model_data_reduced) # 1449 x 50
```

## Plot a correlation matrix of the reduced datasets
```{r}
library(ragg)

# Use ragg PNG device so UTF-8 characters in axis labels render correctly on
# Windows (avoids "conversion failure from UTF-8 to cp1252" errors)
knitr::opts_chunk$set(dev = "ragg_png")

# Plot a correlation matrix of the reduced datasets using ggplot2 and ggcorrplot
# For moss model data (Numerical Features)
moss_num_data <- moss_model_data_reduced |> dplyr::select(where(is.numeric), -richness_moss, -yearMoos)
moss_cor_matrix <- cor(moss_num_data, use = "pairwise.complete.obs")
p_moss_num <- ggcorrplot(moss_cor_matrix, method = "circle", type = "lower", lab = TRUE, title = "Numeric Correlation Matrix (Moss)", ggtheme = ggplot2::theme_minimal(), tl.srt = 90)
print(p_moss_num)
ggsave(file.path(DIR_MDL1_PLOTS, "moss_numeric_cor.png"), plot = p_moss_num, width = 12, height = 10, bg = "white")

# For moss model data (Categorical Features via Dummy-Encoding)
moss_cat_data <- model.matrix(~ 0 + ., data = moss_model_data_reduced |> dplyr::select(where(is.factor)))
moss_cat_cor_matrix <- cor(moss_cat_data, use = "pairwise.complete.obs")
p_moss_cat <- ggcorrplot(moss_cat_cor_matrix, method = "circle", type = "lower", lab = FALSE, title = "Categorical Correlation Matrix (Moss)", ggtheme = ggplot2::theme_minimal(), tl.srt = 90)
print(p_moss_cat)
ggsave(file.path(DIR_MDL1_PLOTS, "moss_categorical_cor.png"), plot = p_moss_cat, width = 16, height = 14, bg = "white")

# For plant model data (Numerical Features)
plant_num_data <- plant_model_data_reduced |> dplyr::select(where(is.numeric), -richness_plant, -yearPl)
plant_cor_matrix <- cor(plant_num_data, use = "pairwise.complete.obs")
p_plant_num <- ggcorrplot(plant_cor_matrix, method = "circle", type = "lower", lab = TRUE, title = "Numeric Correlation Matrix (Plants)", ggtheme = ggplot2::theme_minimal(), tl.srt = 90)
print(p_plant_num)
ggsave(file.path(DIR_MDL1_PLOTS, "plant_numeric_cor.png"), plot = p_plant_num, width = 12, height = 10, bg = "white")

# For plant model data (Categorical Features via Dummy-Encoding)
plant_cat_data <- model.matrix(~ 0 + ., data = plant_model_data_reduced |> dplyr::select(where(is.factor)))
plant_cat_cor_matrix <- cor(plant_cat_data, use = "pairwise.complete.obs")
p_plant_cat <- ggcorrplot(plant_cat_cor_matrix, method = "circle", type = "lower", lab = FALSE, title = "Categorical Correlation Matrix (Plants)", ggtheme = ggplot2::theme_minimal(), tl.srt = 90)
print(p_plant_cat)
ggsave(file.path(DIR_MDL1_PLOTS, "plant_categorical_cor.png"), plot = p_plant_cat, width = 16, height = 14, bg = "white")
```

## Create a DataExplorer report
```{r}
# Use DataExplorer to create a report of the relationship between richness and environmental variables
DataExplorer::create_report(moss_model_data_reduced, y = "richness_moss", output_file = file.path(DIR_MDL1_REPORTS, "moss_model_1_data_report.html"))
DataExplorer::create_report(plant_model_data_reduced, y = "richness_plant", output_file = file.path(DIR_MDL1_REPORTS, "plant_model_1_data_report.html"))
```