# Load libraries
```{r}
library(dplyr)
library(sf)
library(randomForest)
library(ranger)
library(caret)
library(performance)
```

# Load data
```{r}
source("_config.r")

# Load plant model data
plant_data <- readRDS(PATH_MDL1_PLANT_DATA)
plant_richness <- readRDS(PATH_PLANT_RICHNESS)
```

# Data preparation
## Remove target column and plot ID
```{r}
# Remove year and ID columns
plant_data <- plant_data |>
  dplyr::select(-yearPl, -aID_STAO)

# Keep only climate variables for the 15 year aggregate
plant_data <- plant_data |>
  dplyr::select(-dplyr::matches("_(5|10|20|30)yr$"))

dim(plant_data) # 1449 x 120
```

# Species richness modelling Random forest
## Train test split
```{r}
set.seed(123)
train_idx <- caret::createDataPartition(plant_data$richness_plant, p = 0.8, list = FALSE)
plant_train <- plant_data[train_idx, ]
plant_test  <- plant_data[-train_idx, ]
```

## Fit model with hyperparameter tuning, mtry and min.node.size
In this step i try to find the optimal mtry and min.node.size values for the random forest model using caret's train function with repeated cross-validation. 
Mtry is a hyperparameter that often significantly impacts the performance of random forest models, as it controls the number of predictors considered at each split. Min.node.size is another important hyperparameter that determines the minimum number of samples required to split a node, affecting the model's complexity and performance.
I use a repeated cross-validation approach to ensure that the tuning process is robust and not overly influenced by random variations in the data splits.
```{r}
# Define hyper parameters for the random forest model

# Repeated CV for robust tuning
train_control <- caret::trainControl(
  method = "repeatedcv",
  number = 5,          # 5-fold CV
  repeats = 10,        # repeated 10 times
  savePredictions = "final",
  returnResamp = "all"
)

# Tune mtry only, fix trees high for stable importance
plant_grid <- base::expand.grid(
  mtry = c(10, 20, floor(ncol(plant_train) / 3), floor(ncol(plant_train) * 0.75), ncol(plant_train) - 1), # Explore a range of mtry values
  splitrule = "variance",
  min.node.size = c(3, 5, 7, 10) # Explore different node sizes for better performance
)

# Train model with repeated CV
rf_plant_tune <- caret::train(
  richness_plant ~ .,
  set.seed(123), # Set seed for reproducibility
  method = "ranger",
  importance = "none", # We will assess importance later with the final model
  data = plant_train, # Use only training data for tuning
  num.trees = 500,   # fixed high for stable forest
  trControl = train_control,
  tuneGrid = plant_grid,
  na.action = na.omit,
  num.threads = max(1, parallel::detectCores() - 1),
  verbose = TRUE
)

# Save the tuned model
saveRDS(rf_plant_tune, file = file.path(DIR_MDL1_OUT, "rf_plant_tune.rds"))

# Save the mtry ranking table
write.csv(rf_plant_tune$results, file = file.path(DIR_MDL1_OUT, "rf_plant_tune_ranking.csv"), row.names = FALSE)
```

## Assess OOB performance across different numbers of trees
```{r}
# Evaluate OOB performance across different numbers of trees
n_trees_values <- c(5, 20, 50, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)

oob_metrics <- lapply(n_trees_values, function(n_trees) {
  rf_tmp <- ranger::ranger(
    richness_plant ~ .,
    data = plant_train,
    num.trees = n_trees,
    mtry = floor(ncol(plant_train) / 3), # The best mtry from tuning
    min.node.size = 5, # The best min.node.size from tuning
    splitrule = "variance",
    importance = "none",
    num.threads = max(1, parallel::detectCores() - 1),
    seed = 123 # Set seed for reproducibility
  )

  mse_oob <- rf_tmp$prediction.error
  r2_oob <- rf_tmp$r.squared

  data.frame(
    n_trees = n_trees,
    oob_mse = mse_oob,
    oob_r2 = r2_oob
  )
})

# Combine OOB metrics into a single data frame
oob_metrics_df <- dplyr::bind_rows(oob_metrics)

# Save OOB metrics for analysis
write.csv(oob_metrics_df, file = file.path(DIR_MDL1_OUT, "rf_plant_tune_oob_tree_metrics.csv"), row.names = FALSE)
```

### Plot OOB MSE and R2 across different numbers of trees
```{r}
# Calculate scaling parameters for plotting OOB MSE and R2 on the same plot
mse_min <- min(oob_metrics_df$oob_mse, na.rm = TRUE)
mse_max <- max(oob_metrics_df$oob_mse, na.rm = TRUE)
r2_min <- min(oob_metrics_df$oob_r2, na.rm = TRUE)
r2_max <- max(oob_metrics_df$oob_r2, na.rm = TRUE)

oob_metrics_df <- oob_metrics_df |>
  dplyr::mutate(
    oob_r2_scaled = (oob_r2 - r2_min) / (r2_max - r2_min) * (mse_max - mse_min) + mse_min
  )

ggplot(oob_metrics_df, ggplot2::aes(x = n_trees)) +
  geom_line(ggplot2::aes(y = oob_mse, color = "OOB MSE"), linewidth = 1) +
  geom_point(ggplot2::aes(y = oob_mse, color = "OOB MSE"), size = 2) +
  geom_line(ggplot2::aes(y = oob_r2_scaled, color = "OOB R2"), linewidth = 1) +
  geom_point(ggplot2::aes(y = oob_r2_scaled, color = "OOB R2"), size = 2) +
  scale_y_continuous(
    name = "OOB MSE",
    sec.axis = ggplot2::sec_axis(
      ~ (. - mse_min) / (mse_max - mse_min) * (r2_max - r2_min) + r2_min,
      name = "OOB R2"
    )
  ) +
  scale_color_manual(values = c("OOB MSE" = "steelblue", "OOB R2" = "darkgreen")) +
  labs(
    title = "OOB Metrics by Number of Trees",
    x = "Number of Trees",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(aspect.ratio = 0.6)

# Save the OOB metrics plot
ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_plant_tune_oob_tree_metrics.png"), width = 8, height = 6, dpi = 300)
```

## Fit final model with stable settings
After finding the optimal mtry and min.node.size value, I fit the final random forest model using the ranger package with the best mtry and a high number of trees to ensure stability.
```{r}
rf_plant_final <- ranger::ranger(
  richness_plant ~ .,
  data = plant_train, # Use only training data for final model
  num.trees = 500,   # Number of trees for stable prediction
  mtry = floor(ncol(plant_train) / 3), # best mtry from tuning
  min.node.size = 5, # best min.node.size from tuning
  splitrule = "variance",
  importance = "permutation", # Set importance to permutation for final model
  num.threads = max(1, parallel::detectCores() - 1),
  seed = 123 # Set seed for reproducibility
)

rf_plant_final
# R squared OOB: 0.32
# MSE OOB: 153.91

# Save the final model
saveRDS(rf_plant_final, file = file.path(DIR_MDL1_OUT, "rf_plant_final.rds"))
```

### Evaluate on test data
The final step is to evaluate the performance of the fitted model on the test data, which was not used during training or tuning. This provides an unbiased estimate of how well the model generalizes to new, unseen data.
```{r}
# Predict on test data
rf_plant_test_pred <- predict(rf_plant_final, data = plant_test)

# Calculate performance metrics
mse_plant_test <- mean((rf_plant_test_pred$predictions - plant_test$richness_plant)^2)
r2_plant_test <- 1 - sum((rf_plant_test_pred$predictions - plant_test$richness_plant)^2) / sum((plant_test$richness_plant - mean(plant_test$richness_plant))^2)

r2_plant_test
# R squared: 0.37
mse_plant_test 
# MSE: 154.43
```

### Calculate variable importance
Finally, I calculate the variable importance using the permutation method (Altmann's method) to identify which predictors are most influential in predicting plant richness.
```{r}
# Calculate variable importance using permutation method (Altmann's method)
importance_pvalues_plant_obj <- ranger::importance_pvalues(
  rf_plant_final,
  method = "altmann",
  num.permutations = 100,
  formula = richness_plant ~ .,
  data = plant_train, # Use training data for importance calculation to avoid data leakage
  num.threads = parallel::detectCores() - 1,
  seed = 123 # Set seed for reproducibility
)

# Convert importance results to a data frame for easier handling
importance_pvalues_plant_df <- data.frame(
  Variable = rownames(importance_pvalues_plant_obj),
  importance_pvalues_plant_obj
) |>
  dplyr::arrange(pvalue)

# Save the final model and importance results
write.csv(importance_pvalues_plant_df, file = file.path(DIR_MDL1_OUT, "rf_plant_imp_pval.csv"), row.names = FALSE)
```

#### Plot variable importance of the 20 most important variables
```{r}
# Plot variable importance of the 20 most important variables
top_20_importance <- importance_pvalues_plant_df |>
  slice_head(n = 20)

# Create a bar plot of the top 20 variable importance
ggplot(top_20_importance, aes(x = reorder(Variable, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0("p=", round(pvalue, 3))), hjust = -0.1, size = 3) +
  coord_flip() +
  expand_limits(y = max(top_20_importance$importance) * 1.2) +
  labs(
    title = "Top 20 Variable Importance",
    x = "Variable",
    y = "Importance"
  ) +
  theme_minimal()

# Save the variable importance plot
ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_plant_imp_top20.png"), width = 8, height = 6, dpi = 300)
```

#### Partial dependence plots for top 20 variables
```{r}
library(pdp)

# pred.fun must return a single mean value per grid point, not a vector
ranger_pred <- function(object, newdata) mean(predict(object, data = newdata)$predictions)

# Compute marginal PDP for each of the top 20 variables and combine into one data frame
top_20_vars <- importance_pvalues_plant_df |>
  slice_head(n = 20) |>
  pull(Variable)

pdp_top20_df <- dplyr::bind_rows(lapply(top_20_vars, function(var) {
  pd <- pdp::partial(
    rf_plant_final,
    pred.var        = var,
    train           = plant_train,
    pred.fun        = ranger_pred,
    grid.resolution = 40   # evaluate at 20 evenly-spaced points — avoids jagged lines from many unique values
  )
  x_val <- pd[[var]]
  data.frame(
    variable    = var,
    x_value     = as.character(x_val),   # keep as character; convert to numeric later per type
    x_is_factor = is.factor(x_val) || is.character(plant_train[[var]]) || is.factor(plant_train[[var]]),
    yhat        = pd$yhat
  )
}))

# Order panels by importance rank
pdp_top20_df$variable <- factor(pdp_top20_df$variable, levels = top_20_vars)

# Split into numeric and categorical subsets for appropriate geoms
pdp_numeric <- pdp_top20_df |>
  dplyr::filter(!x_is_factor) |>
  dplyr::mutate(x_value = as.numeric(x_value))

pdp_factor <- pdp_top20_df |>
  dplyr::filter(x_is_factor)

# Numeric variables: line plot
p_num <- ggplot(pdp_numeric, aes(x = x_value, y = yhat)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(title = "Partial Dependence — Numeric Variables", x = "Variable Value", y = "Predicted Plant Richness") +
  theme_minimal(base_size = 9)

# Categorical variables: bar plot
p_cat <- ggplot(pdp_factor, aes(x = x_value, y = yhat)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(title = "Partial Dependence — Categorical Variables", x = "Category", y = "Predicted Plant Richness") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_num
p_cat

ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_plant_pdp_top20_numeric.png"), plot = p_num, width = 14, height = 12, dpi = 300)
ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_plant_pdp_top20_categorical.png"), plot = p_cat, width = 14, height = 8, dpi = 300)
```

#### Interaction partial dependence: AS_4 x LU_10
```{r}
# 2D interaction PDP — heatmap of predicted richness across the AS_4 x LU_10 value space
pdp_as4_lu10 <- pdp::partial(
  rf_plant_final,
  pred.var = c("AS_4", "LU_10"),
  train    = plant_train,
  pred.fun = ranger_pred
)

ggplot(pdp_as4_lu10, aes(x = AS_4, y = LU_10, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted\nRichness") +
  labs(
    title = "Interaction Partial Dependence: AS_4 x LU_10",
    x = "AS_4",
    y = "LU_10"
  ) +
  theme_minimal()

ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_plant_pdp_AS_4_LU_10.png"), width = 8, height = 6, dpi = 300)
```

## Fit model using cv
I now fit the model using caret's train function with the same repeated cross-validation setup to get performance estimates across all folds and repeats, which provides a more robust assessment of model performance than just looking at OOB metrics from the final model.
```{r}
set.seed(123)
rf_plant_cv <- caret::train(
  richness_plant ~ .,
  data        = plant_data,
  method      = "ranger",
  importance  = "none",
  num.trees   = 500,
  trControl   = train_control,
  tuneGrid    = expand.grid(
    mtry          = floor(ncol(plant_data) / 3),
    splitrule     = "variance",
    min.node.size = 5
  ),
  na.action   = na.omit,
  num.threads = max(1, parallel::detectCores() - 1),
  verbose     = TRUE
)

rf_plant_cv
rf_plant_cv$results
rf_plant_cv$resample

# Manual summary across all folds/repeats
resamp <- rf_plant_cv$resample
data.frame(
  metric = c("RMSE", "Rsquared", "MAE"),
  mean   = colMeans(resamp[, c("RMSE", "Rsquared", "MAE")], na.rm = TRUE),
  sd     = apply(resamp[, c("RMSE", "Rsquared", "MAE")], 2, sd, na.rm = TRUE)
)

# Save the CV model and results
saveRDS(rf_plant_cv, file = file.path(DIR_MDL1_OUT, "rf_plant_full_cv.rds"))
write.csv(rf_plant_cv$results, file = file.path(DIR_MDL1_OUT, "rf_plant_full_cv_results.csv"), row.names = FALSE)
```

## Fit single model on full dataset with best hyperparameters
Finally, I fit a single random forest model on the entire dataset using the best mtry and min.node.size values found during tuning. This model can be used for final variable importance assessment and partial dependence analysis, as it is trained on all available data.
```{r}
rf_plant_final_full <- ranger::ranger(
  richness_plant ~ .,
  data          = plant_data,      # full dataset
  num.trees     = 500,
  mtry          = rf_plant_cv$bestTune$mtry,
  min.node.size = rf_plant_cv$bestTune$min.node.size,
  splitrule     = "variance",
  importance    = "permutation",
  seed          = 123
)

# Save the final full model
saveRDS(rf_plant_final_full, file = file.path(DIR_MDL1_OUT, "rf_plant_final_full.rds"))

importance_pvalues_plant_obj <- ranger::importance_pvalues(
  rf_plant_final_full,
  method           = "altmann",
  num.permutations = 100,
  formula          = richness_plant ~ .,
  data             = plant_data,   # consistent: model was trained on full data
  num.threads      = parallel::detectCores() - 1,
  seed             = 123
)

# Save the importance results for the full model
importance_pvalues_plant_df <- data.frame(
  Variable = rownames(importance_pvalues_plant_obj),
  importance_pvalues_plant_obj
) |>
  dplyr::arrange(pvalue)

importance_pvalues_plant_df

write.csv(importance_pvalues_plant_df, file = file.path(DIR_MDL1_OUT, "rf_plant_full_imp_pval.csv"), row.names = FALSE)
```
