# Load libraries
```{r}
library(dplyr)
library(sf)
library(randomForest)
library(ranger)
library(readxl)
library(caret)
library(performance)
library(pdp)
library(ggplot2)
library(patchwork)
library(vegan)
```

# Load data
```{r}
source("_config.r")

# Load moss model data
moss_data <- readRDS(PATH_MDL3_MOSS_DATA)
moss_richness <- readRDS(PATH_MOSS_RICHNESS)
```

# Data preparation
## Remove target column and plot ID
```{r}
# Remove year and ID columns
moss_data <- moss_data |>
  dplyr::select(-yearMoos, -aID_STAO)

dim(moss_data) # 1435 x 21

# Print variable names for reference
colnames(moss_data)
```

# Species richness modelling Random forest
## Stable CV modelling to tune parameter
Here i do CV tuning using the entire dataset since the samples are limited and the initial selection strongly affects the result. Thus the standard Train-Test Split procedure is not implemented.
```{r}
# Set repeated CV for robust tuning
train_control <- caret::trainControl(
  method = "repeatedcv",
  number = 5,          # 5-fold CV
  repeats = 10,        # repeated 10 times
  savePredictions = "all",
  returnResamp = "all"
)

# Tune mtry only, fix trees high for stable importance
moss_grid <- base::expand.grid(
  mtry = c(floor(ncol(moss_data) / 3), floor(ncol(moss_data) * 0.75), ncol(moss_data) - 1), # Explore a range of mtry values
  splitrule = "variance",
  min.node.size = c(3, 5, 7, 10) # Explore different node sizes for better performance
)

set.seed(123)
rf_moss_cv_tune <- caret::train(
  richness_moss ~ .,
  set.seed(123),
  data        = moss_data, # Use full data for CV modeling
  method      = "ranger",
  num.trees   = 500,
  trControl   = train_control,
  tuneGrid    = moss_grid,
  na.action   = na.omit,
  num.threads = max(1, parallel::detectCores() - 1),
  verbose     = TRUE
)

rf_moss_cv_tune

# Save the CV model and results
saveRDS(rf_moss_cv_tune, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_full_cv_tune.rds"))
write.csv(rf_moss_cv_tune$results, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_full_cv_tune_results.csv"), row.names = FALSE)
```

## CV model using best hyperparameters
The best hyperparameters are essentially the default ones, thus the CV model is run again with the standard parameters to get a more stable estimate of performance across all folds and repeats.
```{r}
# Set best hyperparameters
best_mtry <- floor(sqrt(ncol(moss_data)))
best_min_node_size <- 3

# Fit CV model with best hyperparameters to get performance metrics across all folds/repeats
set.seed(123)
rf_moss_cv_std <- caret::train(
  richness_moss ~ .,
  data        = moss_data,
  method      = "ranger",
  num.trees   = 500,
  trControl   = train_control,
  tuneGrid    = expand.grid(
    mtry          = best_mtry,
    splitrule     = "variance",
    min.node.size = best_min_node_size
  ),
  na.action   = na.omit,
  num.threads = max(1, parallel::detectCores() - 1),
  verbose     = TRUE
)

rf_moss_cv_std

# Manual summary across all folds/repeats with standard parameters
resamp_std <- rf_moss_cv_std$resample
data.frame(
  metric = c("RMSE", "Rsquared", "MAE"),
  mean   = colMeans(resamp_std[, c("RMSE", "Rsquared", "MAE")], na.rm = TRUE),
  sd     = apply(resamp_std[, c("RMSE", "Rsquared", "MAE")], 2, sd, na.rm = TRUE)
)
#           metric      mean         sd
# RMSE         RMSE 6.4478985 0.35236466
# Rsquared Rsquared 0.4705521 0.03723306
# MAE           MAE 4.7641125 0.20857580

# Save the CV model and results
saveRDS(rf_moss_cv_std, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_std_cv.rds"))
write.csv(rf_moss_cv_std$results, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_std_cv_results.csv"), row.names = FALSE)
```

## Assess performance across n-trees using best hyperparameters
```{r}
# Assess performance across n-trees using best hyperparameters
n_trees_values <- c(5, 20, 50, 100, 200, 400, 600, 800)

# Loop over n-trees values and collect CV metrics
cv_tree_metrics <- lapply(n_trees_values, function(n_trees) {
  rf_tmp <- caret::train(
    richness_moss ~ .,
    data      = moss_data,
    method    = "ranger",
    importance = "none",
    num.trees = n_trees,
    trControl = train_control,
    tuneGrid  = expand.grid(
      mtry          = best_mtry,
      splitrule     = "variance",
      min.node.size = best_min_node_size
    ),
    na.action   = na.omit,
    num.threads = max(1, parallel::detectCores() - 1),
    seed        = 123
  )

  resamp <- rf_tmp$resample
  data.frame(
    n_trees      = n_trees,
    RMSE_mean    = mean(resamp$RMSE,      na.rm = TRUE),
    RMSE_sd      = sd(resamp$RMSE,        na.rm = TRUE),
    Rsquared_mean = mean(resamp$Rsquared, na.rm = TRUE),
    Rsquared_sd  = sd(resamp$Rsquared,    na.rm = TRUE),
    MAE_mean     = mean(resamp$MAE,       na.rm = TRUE),
    MAE_sd       = sd(resamp$MAE,         na.rm = TRUE)
  )
})

# Combine CV metrics into a single data frame
cv_tree_metrics_df <- dplyr::bind_rows(cv_tree_metrics)
print(cv_tree_metrics_df)

# Save CV tree metrics for analysis
write.csv(cv_tree_metrics_df, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_tune_cv_tree_metrics.csv"), row.names = FALSE)

# Plot the performance metrics across n-trees
cv_tree_metrics_long <- cv_tree_metrics_df |>
  tidyr::pivot_longer(cols = -n_trees, names_to = c("metric", "stat"), names_sep = "_") |>
  tidyr::pivot_wider(names_from = stat, values_from = value)

ggplot(cv_tree_metrics_long, aes(x = n_trees, y = mean)) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw(base_size = 9) +
  labs(x = "Number of Trees", y = "Performance Metric")

ggsave(
  filename = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_cv_tree_metrics.png"),
  plot     = last_plot(),
  width    = 6,
  height   = 4,
  dpi      = 300,
  bg       = "white"
)
```

# Fit final model using all data
```{r}
# Fit final model using all data with best hyperparameters
rf_moss_final <- ranger::ranger(
  formula     = richness_moss ~ .,
  data        = moss_data,
  num.trees   = 500,
  mtry        = best_mtry,
  min.node.size = best_min_node_size,
  importance   = "permutation",
  num.threads = max(1, parallel::detectCores() - 1),
  write.forest = TRUE,
  keep.inbag  = TRUE,
  oob.error   = TRUE,
  verbose     = TRUE,
  seed        = 123
)

rf_moss_final
# OOB prediction error (MSE):       41.20176 
# R squared (OOB):                  0.4702841 

# Save final model
saveRDS(rf_moss_final, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_final_model.rds"))
```

## Variable importance
```{r}
# Extract variable importance p-values via Altmann permutation test
set.seed(123)
imp_pvals <- ranger::importance_pvalues(
  rf_moss_final,
  method            = "altmann",
  formula           = richness_moss ~ .,
  data              = moss_data,
  num.permutations  = 100,
  num.threads      = max(1, parallel::detectCores() - 1),
  verbose           = TRUE
)

var_importance <- as.data.frame(imp_pvals) |>
  tibble::rownames_to_column("variable") |>
  dplyr::arrange(pvalue)

var_importance

# Save variable importance
write.csv(var_importance, file = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_variable_importance.csv"), row.names = FALSE)
```

## Morans I
```{r}
library(spdep)

# Load final model and training data (with aID_STAO still present for coordinate join)
rf_moss_final <- readRDS(file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_final_model.rds"))
moss_data_full <- readRDS(PATH_MDL3_MOSS_DATA) |>
  dplyr::select(-yearMoos)  # keep aID_STAO for joining coordinates

# Load plot coordinates
plot_coords <- readRDS(PATH_PLOT_COORDS_RDS) |>
  dplyr::select(aID_STAO, XKoord_LV95, YKoord_LV95)

# OOB predictions are aligned to moss_data_full row order
# (ranger stores OOB predictions for every training observation)
moss_data_full$oob_pred <- rf_moss_final$predictions
moss_data_full$residual <- moss_data_full$richness_moss - moss_data_full$oob_pred

# Join coordinates
resid_coords <- moss_data_full |>
  dplyr::left_join(plot_coords, by = "aID_STAO")

cat(sprintf("Observations used for Moran's I: %d\n", nrow(resid_coords)))

# Build spatial weights: inverse-distance weighted neighbours within 50 km
# (50 km captures the nearest BDM plots across the Swiss grid)
coords_mat <- as.matrix(resid_coords[, c("XKoord_LV95", "YKoord_LV95")])
nb  <- spdep::dnearneigh(coords_mat, d1 = 0, d2 = 6000)  # 50 km threshold
wts <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)

# Moran's I test on OOB residuals
moran_result <- spdep::moran.test(
  resid_coords$residual,
  listw       = wts,
  zero.policy = TRUE,
  alternative = "two.sided"
)
moran_result
# Moran I statistic close to 0 and p > 0.05 → no meaningful spatial autocorrelation
# Moran I statistic significantly > 0      → positive spatial autocorrelation (nearby plots share unexplained variance)

# Save result summary
moran_summary <- data.frame(
  statistic = moran_result$statistic,
  expected  = moran_result$estimate["Expected"],
  variance  = moran_result$estimate["Variance"],
  p_value   = moran_result$p.value
)
write.csv(moran_summary,
          file      = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_morans_i.csv"),
          row.names = FALSE)

# Spatial plot of OOB residuals for visual inspection
ggplot(resid_coords, aes(x = XKoord_LV95, y = YKoord_LV95, color = residual)) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_gradient2(low = "#4DBBD5", mid = "white", high = "#E64B35",
                        midpoint = 0, name = "OOB residual") +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Spatial distribution of OOB residuals — moss richness RF",
       x = "X (LV95)", y = "Y (LV95)")

ggsave(
  filename = file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_residuals_spatial.png"),
  plot     = last_plot(),
  width    = 10, height = 6, dpi = 300, bg = "white"
)
```