# Load libraries
```{r}
library(dplyr)
library(sf)
library(randomForest)
library(ranger)
library(caret)
library(performance)
```

# Load data
```{r}
source("_config.r")

# Load moss model data
moss_data <- readRDS(PATH_MDL1_MOSS_DATA)
moss_richness <- readRDS(PATH_MOSS_RICHNESS)
```

# Data preparation
## Remove target column and plot ID
```{r}
# Remove year and ID columns
moss_data <- moss_data |>
  dplyr::select(-yearMoos, -aID_STAO)

# Keep only climate variables for the 15 year aggregate
moss_data <- moss_data |>
  dplyr::select(-dplyr::matches("_(5|10|20|30)yr$"))

dim(moss_data) # 1442 x 120
```

# Species richness modelling Random forest
## Train test split
```{r}
set.seed(123)
train_idx <- caret::createDataPartition(moss_data$richness_moss, p = 0.8, list = FALSE)
moss_train <- moss_data[train_idx, ]
moss_test  <- moss_data[-train_idx, ]
```

## Assess OOB performance across different numbers of trees
```{r}
# Evaluate OOB performance across different numbers of trees
n_trees_values <- c(5, 20, 50, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)

oob_metrics <- lapply(n_trees_values, function(n_trees) {
  rf_tmp <- ranger::ranger(
    richness_moss ~ .,
    data = moss_train,
    num.trees = n_trees,
    mtry = floor(ncol(moss_train) / 3),
    min.node.size = 5,
    splitrule = "variance",
    importance = "none",
    num.threads = max(1, parallel::detectCores() - 1),
    seed = 123
  )

  data.frame(
    n_trees = n_trees,
    oob_mse = rf_tmp$prediction.error,
    oob_r2  = rf_tmp$r.squared
  )
})

# Combine OOB metrics into a single data frame
oob_metrics_df <- dplyr::bind_rows(oob_metrics)

# Save OOB metrics for analysis
write.csv(oob_metrics_df, file = file.path(DIR_MDL1_OUT, "rf_moss_tune_oob_tree_metrics.csv"), row.names = FALSE)
```

### Plot OOB MSE and R2 across different numbers of trees
```{r}
# Calculate scaling parameters for plotting OOB MSE and R2 on the same plot
mse_min <- min(oob_metrics_df$oob_mse, na.rm = TRUE)
mse_max <- max(oob_metrics_df$oob_mse, na.rm = TRUE)
r2_min  <- min(oob_metrics_df$oob_r2, na.rm = TRUE)
r2_max  <- max(oob_metrics_df$oob_r2, na.rm = TRUE)

oob_metrics_df <- oob_metrics_df |>
  dplyr::mutate(
    oob_r2_scaled = (oob_r2 - r2_min) / (r2_max - r2_min) * (mse_max - mse_min) + mse_min
  )

ggplot(oob_metrics_df, ggplot2::aes(x = n_trees)) +
  geom_line(ggplot2::aes(y = oob_mse, color = "OOB MSE"), linewidth = 1) +
  geom_point(ggplot2::aes(y = oob_mse, color = "OOB MSE"), size = 2) +
  geom_line(ggplot2::aes(y = oob_r2_scaled, color = "OOB R2"), linewidth = 1) +
  geom_point(ggplot2::aes(y = oob_r2_scaled, color = "OOB R2"), size = 2) +
  scale_y_continuous(
    name = "OOB MSE",
    sec.axis = ggplot2::sec_axis(
      ~ (. - mse_min) / (mse_max - mse_min) * (r2_max - r2_min) + r2_min,
      name = "OOB R2"
    )
  ) +
  scale_color_manual(values = c("OOB MSE" = "steelblue", "OOB R2" = "darkgreen")) +
  labs(
    title = "OOB Metrics by Number of Trees",
    x = "Number of Trees",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(aspect.ratio = 0.6)

ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_moss_tune_oob_tree_metrics.png"), width = 8, height = 6, dpi = 300)
```

## Fit final model with stable settings
```{r}
rf_moss_final <- ranger::ranger(
  richness_moss ~ .,
  data = moss_train, # Use only training data for final model
  num.trees = 500, # Number of trees for stable prediction
  mtry = floor(ncol(moss_train) / 3),
  min.node.size = 5,
  splitrule = "variance",
  importance = "permutation",
  num.threads = max(1, parallel::detectCores() - 1),
  seed = 123
)

rf_moss_final
# R squared OOB: 0.47
# MSE OOB: 40.67

saveRDS(rf_moss_final, file = file.path(DIR_MDL1_OUT, "rf_moss_final.rds"))
```

### Evaluate on test data
```{r}
# Predict on test data
rf_moss_test_pred <- predict(rf_moss_final, data = moss_test)

# Calculate performance metrics
mse_moss_test <- mean((rf_moss_test_pred$predictions - moss_test$richness_moss)^2)
r2_moss_test  <- 1 - sum((rf_moss_test_pred$predictions - moss_test$richness_moss)^2) /
                     sum((moss_test$richness_moss - mean(moss_test$richness_moss))^2)

r2_moss_test
# R squared: 0.38
mse_moss_test
# MSE: 50.64
```

### Calculate variable importance
```{r}
# Calculate variable importance using permutation method (Altmann's method)
importance_pvalues_moss_obj <- ranger::importance_pvalues(
  rf_moss_final,
  method = "altmann",
  num.permutations = 100,
  formula = richness_moss ~ .,
  data = moss_train, # Use training data for importance calculation to avoid data leakage
  num.threads = parallel::detectCores() - 1,
  seed = 123 # Set seed for reproducibility
)

# Convert importance results to a data frame for easier handling
importance_pvalues_moss_df <- data.frame(
  Variable = rownames(importance_pvalues_moss_obj),
  importance_pvalues_moss_obj
) |>
  dplyr::arrange(pvalue)

# Save the final model and importance results
write.csv(importance_pvalues_moss_df, file = file.path(DIR_MDL1_OUT, "rf_moss_imp_pval.csv"), row.names = FALSE)
```

#### Plot variable importance of the 20 most important variables
```{r}
# Keep only significant variables (p < 0.05), ordered by importance
sig_importance <- importance_pvalues_moss_df |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::arrange(dplyr::desc(importance))

# Create a bar plot of significant variables
ggplot(sig_importance, aes(x = reorder(Variable, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0("p=", round(pvalue, 3))), hjust = -0.1, size = 3) +
  coord_flip() +
  expand_limits(y = max(sig_importance$importance) * 1.2) +
  labs(
    title = paste0("Significant Variable Importance (p < 0.05, n=", nrow(sig_importance), ")"),
    x = "Variable",
    y = "Importance"
  ) +
  theme_minimal()

# Save the variable importance plot
ggsave(filename = file.path(DIR_MDL1_PLOTS, "rf_moss_imp_significant.png"), width = 8, height = max(4, nrow(sig_importance) * 0.3), dpi = 300)
```

#### Plot the partial dependence of the most important variable
```{r}


```