# Attempt 3, variable importance barplots
## Load libraries
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(pdp)
library(ranger)
library(ggsci)
library(extrafont)
```

## Global plotting settings
```{r}
source("./_config.r")
DIR_PLOTS <- "08_plotting/plots"

# Report column width in inches (147 mm)
REPORT_WIDTH_IN <- 147 / 25.4

# Fixed height per bar in inches with total panel height = n_bars * BAR_HEIGHT + PANEL_OVERHEAD
BAR_HEIGHT    <- 0.09
PANEL_OVERHEAD <- 0.5  # space for axis, margins
LEGEND_HEIGHT  <- 0.6  # extra space at bottom for collected legend
COMBINED_PADDING <- 0.45  # extra margin for top/bottom clipping in combined figures

# Panel height in inches for a given importance data frame
imp_plot_height <- function(imp_df, sig_threshold = 0.05) {
  # Sums up the number of significant variables
  n <- sum(imp_df$pvalue < sig_threshold)
  # Returns the height of the panel in inches, with a minimum height of 0.2 inches
  max(0.2, n * BAR_HEIGHT + PANEL_OVERHEAD)
}

# Load color PALette
PAL <- pal_npg("nrc")(10)

# Set global theme for ggplot2
theme_set(
    theme_classic() + # Start with a base theme
    theme(
    # 1. Globaly set font family and size for all text elements
    text = element_text(
        family = "Sans", # Global font family
        size = 20          # Global base font size
    ),

    # 2. Fine-tuning for specific elements (overrides the global setting)
    # Axis titles
    axis.title = element_text(
        size = 8, 
        color = "black"
    ),

    # Axis text (text on the axes)
    axis.text = element_text(size = 8, color = "black"),

    # Legend titles
    legend.title = element_text(
        size = 8, 
        face = "bold"
    ),
    # Legend text (text next to the symbols in the legend)
    legend.text = element_text(size = 8)
    )
)
```

## Global constants for plotting
```{r}
# Categorical variables — rendered as bar charts instead of line PDPs
CATEGORICAL_VARS <- c("aspect_factor", "mowing_intensity_factor", "mow_int_cat",
                      "TypoCH_1st_code", "LU_4")

# Human-readable labels for every potential predictor (used in PDP and importance plots)
VAR_LABELS <- c(
  # Climate — cross-sectional
  bio3                = "Isothermality (BIO3)",
  bio4                = "Temp. Seasonality (BIO4)",
  bio10               = "Temp. Warmest Quarter (BIO10)",
  bio11               = "Temp. Coldest Quarter (BIO11)",
  bio18               = "Precip. Warmest Quarter (BIO18)",
  bio19               = "Precip. Coldest Quarter (BIO19)",
  radiation_y_15yr    = "Solar Radiation (15-yr avg)",
  # Climate — change baselines
  bio3_baseline       = "Isothermality baseline (BIO3)",
  bio4_baseline       = "Temp. Seasonality baseline (BIO4)",
  bio10_baseline      = "Temp. Warmest Qtr. baseline (BIO10)",
  bio11_baseline      = "Temp. Coldest Qtr. baseline (BIO11)",
  bio18_baseline      = "Precip. Warmest Qtr. baseline (BIO18)",
  bio19_baseline      = "Precip. Coldest Qtr. baseline (BIO19)",
  # Climate — change trends
  bio3_slope          = "Isothermality trend (BIO3)",
  bio4_slope          = "Temp. Seasonality trend (BIO4)",
  bio10_slope         = "Temp. Warmest Qtr. trend (BIO10)",
  bio11_slope         = "Temp. Coldest Qtr. trend (BIO11)",
  bio18_slope         = "Precip. Warmest Qtr. trend (BIO18)",
  bio19_slope         = "Precip. Coldest Qtr. trend (BIO19)",
  # Vegetation/Habitat
  veg_height          = "Vegetation Height (m)",
  TypoCH_1st_code     = "Habitat Type (TypoCH)",
  # Topography/Soil
  aspect_factor       = "Aspect",
  slope               = "Slope (\u00b0)",
  tpi                 = "Topographic Position Index (TPI)",
  twi                 = "Topographic Wetness Index (TWI)",
  soc_stock_0_120     = "Soil Organic Carbon (t/ha)",
  humus_0_30          = "Humus Content",
  kak_0_30            = "Cation Exchange Capacity (KAK)",
  ph_0_30             = "Soil pH",
  sand_0_30           = "Sand Content (%)",
  # Land Use
  LU_4                = "Land Use (4 classes)",
  mow_int_cat         = "Mowing Intensity"
)

# Return a pretty label for a variable name, falling back to cleaned snake_case
var_label <- function(v) {
  if (v %in% names(VAR_LABELS)) VAR_LABELS[[v]] else gsub("_", " ", v)
}

# Variable group membership (covers both cross-sectional and change models)
IMP_VAR_GROUPS <- list(
  "Climate" = c(
    "bio3", "bio4", "bio10", "bio11", "bio18", "bio19", "radiation_y_15yr",
    "bio3_baseline", "bio4_baseline", "bio10_baseline",
    "bio11_baseline", "bio18_baseline", "bio19_baseline",
    "bio3_slope", "bio4_slope", "bio10_slope",
    "bio11_slope", "bio18_slope", "bio19_slope"
  ),
  "Vegetation/Habitat" = c("veg_height", "TypoCH_1st_code"),
  "Topography/Soil" = c(
    "aspect_factor", "slope", "tpi", "twi",
    "soc_stock_0_120", "humus_0_30", "kak_0_30", "ph_0_30", "sand_0_30"
  ),
  "Land Use" = c("LU_4", "mow_int_cat")
)

imp_get_group <- function(var) {
  for (grp in names(IMP_VAR_GROUPS)) {
    if (var %in% IMP_VAR_GROUPS[[grp]]) return(grp)
  }
  return("Other")
}

# Group colours from the global NPG PALette
IMP_GROUP_COLORS <- c(
  "Climate"            = PAL[1],
  "Vegetation/Habitat" = PAL[3],
  "Topography/Soil"    = PAL[9],
  "Land Use"           = PAL[4],
  "Other"              = "gray70"
)

# Significant continuous variables top_n by importance, categoricals excluded
select_cont_vars <- function(imp_df, top_n = 4) {
  imp_df |>
    # Filter for significant continuous variables
    dplyr::filter(pvalue < 0.05, !variable %in% CATEGORICAL_VARS) |>
    # Select top_n by importance
    dplyr::arrange(dplyr::desc(importance)) |>
    dplyr::slice_head(n = top_n) |>
    dplyr::pull(variable)
}

# Compact p-value label (3 significant figures)
fmt_pval <- function(p) {
  ifelse(p < 0.001, "p<0.001", sprintf("p=%.3g", p))
}
```


## Load models and data
```{r}
# Load moss richness model and data
rf_moss_richness   <- readRDS(file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_final_model.rds"))
moss_richness_data <- readRDS(PATH_MDL3_MOSS_DATA) |> dplyr::select(-yearMoos, -aID_STAO)
imp_moss_richness  <- read.csv(file.path(DIR_MDL3_OUT, "moss_richness/rf_moss_variable_importance.csv"))

# Load moss change model and data
rf_moss_change   <- readRDS(file.path(DIR_MDL3_OUT, "moss_change/rf_moss_change_final_model.rds"))
moss_change_data <- readRDS(PATH_MDL3_SLOPES_MOSS) |> dplyr::select(-aID_STAO, -total_change_pct)
imp_moss_change  <- read.csv(file.path(DIR_MDL3_OUT, "moss_change/rf_moss_change_variable_importance.csv"))

# Load plant richness model and data
rf_plant_richness   <- readRDS(file.path(DIR_MDL3_OUT, "plant_richness/rf_plant_final_model.rds"))
plant_richness_data <- readRDS(PATH_MDL3_PLANT_DATA) |> dplyr::select(-yearPl, -aID_STAO)
imp_plant_richness  <- read.csv(file.path(DIR_MDL3_OUT, "plant_richness/rf_plant_variable_importance.csv"))

# Load plant change model and data
rf_plant_change   <- readRDS(file.path(DIR_MDL3_OUT, "plant_change/rf_plant_change_final_model.rds"))
plant_change_data <- readRDS(PATH_MDL3_SLOPES_PLANT) |> dplyr::select(-aID_STAO, -total_change_pct)
imp_plant_change  <- read.csv(file.path(DIR_MDL3_OUT, "plant_change/rf_plant_change_variable_importance.csv"))
```

## Barplot of significant variable importance

Horizontal bar charts show all predictors with permutation importance p < 0.05 per model, sorted by importance. Bars are coloured by environmental group, the p-value from the permutation test is displayed at the end of each bar.

### Main function to build a horizontal barplot of variable importance
```{r}
# Build a sorted horizontal barplot for one model
make_imp_barplot <- function(imp_df, sig_threshold = 0.05, x_max = NULL) {

  # Filter for significant variables, add labels, and sort by importance
  d <- imp_df |>

    # Grabs the significant variables and adds the group and labels for plotting
    dplyr::filter(pvalue < sig_threshold) |>
    dplyr::mutate(

      # Add group, labels and format the p-value for plotting
      group    = sapply(variable, imp_get_group),
      var_lbl  = sapply(variable, var_label),
      pval_lbl = fmt_pval(pvalue)
    ) |>

    # Aranges selected variables by importance and sets the factor levels for plotting
    dplyr::arrange(importance) |>
    dplyr::mutate(var_lbl = factor(var_lbl, levels = unique(var_lbl)))

  # Build the horizontal barplot
  ggplot(d, aes(x = importance, y = var_lbl, fill = group)) +

    # Add bars and p-value labels
    geom_col(width = 0.7) +
    geom_text(aes(label = pval_lbl),
              hjust = -0.08, size = 2.8, color = "gray40") +

    # Set the fill colors, remove legend title, and order the legend
    scale_fill_manual(
      values = IMP_GROUP_COLORS,
      name   = NULL,
    ) +

    # Set the x-axis limits and expand the right side to make room for p-value labels
    # Can be passed as an argument to adjust for different importance ranges across models
    scale_x_continuous(
      expand = expansion(mult = c(0, 0), add = c(0, 0)),
      limits = if (!is.null(x_max)) c(0, x_max) else NULL
    ) +
    labs(x = NULL, y = NULL) +
    theme(
      # Remove the legend for individual plots since it will be collected in combined plots
      legend.position = "none"
    )
}
```

### Moss richness
```{r}
p_imp_moss_richness <- make_imp_barplot(imp_moss_richness, x_max = 20)

ggsave(
  filename = file.path(DIR_PLOTS, "moss_richness/rf_moss_importance_barplot.png"),
  plot     = p_imp_moss_richness,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_moss_richness),
  dpi      = 300, bg = "white"
)
```

### Plant richness
```{r}
p_imp_plant_richness <- make_imp_barplot(imp_plant_richness, x_max = 90)

ggsave(
  filename = file.path(DIR_PLOTS, "plant_richness/rf_plant_importance_barplot.png"),
  plot     = p_imp_plant_richness,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_plant_richness),
  dpi      = 300, bg = "white"
)
```

### Moss change (richness slope)
```{r}
p_imp_moss_change <- make_imp_barplot(imp_moss_change, x_max = 0.017)

ggsave(
  filename = file.path(DIR_PLOTS, "moss_change/rf_moss_change_importance_barplot.png"),
  plot     = p_imp_moss_change,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_moss_change),
  dpi      = 300, bg = "white"
)
```

### Plant change (richness slope)
```{r}
p_imp_plant_change <- make_imp_barplot(imp_plant_change, x_max = 0.115)

ggsave(
  filename = file.path(DIR_PLOTS, "plant_change/rf_plant_change_importance_barplot.png"),
  plot     = p_imp_plant_change,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_plant_change),
  dpi      = 300, bg = "white"
)
```

### Combined richness importance figure (moss + plant)
```{r}
combined_imp_richness <- (
  # Combine moss and plant richness importance plots into a single figure
  p_imp_moss_richness /
  (p_imp_plant_richness + labs(x = "Permutation Importance (% decrease in explained R\u00b2)"))
) +
  plot_layout(
    # Heights proportional to bar count so each bar is the same physical height
    guides  = "collect",
    heights = c(
      sum(imp_moss_richness$pvalue  < 0.05),
      sum(imp_plant_richness$pvalue < 0.05)
    )
  ) &
  theme(legend.position = "bottom", legend.key.size = unit(4.5, "mm"), legend.spacing.x = unit(10, "mm"),
        legend.box.margin = margin(t = -2, unit = "mm"))

ggsave(
  filename = file.path(DIR_PLOTS, "combined_importance_richness_barplot.png"),
  plot     = combined_imp_richness,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_moss_richness) + imp_plot_height(imp_plant_richness) + LEGEND_HEIGHT + COMBINED_PADDING,
  dpi      = 300, bg = "white"
)
```

### Combined change importance figure (moss + plant)
```{r}
combined_imp_change <- (
  # Combine moss and plant change importance plots into a single figure
  p_imp_moss_change /
  (p_imp_plant_change +
     labs(x = "Permutation Importance (% decrease in explained R\u00b2)") +
     theme(
       legend.position   = "bottom",
       legend.key.size   = unit(4.5, "mm"),
       legend.spacing.x  = unit(10, "mm"),
       legend.box.margin = margin(t = -2, unit = "mm")
     ))
) +
  plot_layout(
    # Heights proportional to bar count so each bar is the same physical height
    heights = c(
      sum(imp_moss_change$pvalue  < 0.05),
      sum(imp_plant_change$pvalue < 0.05)
    )
  )

ggsave(
  filename = file.path(DIR_PLOTS, "combined_importance_change_barplot.png"),
  plot     = combined_imp_change,
  width    = REPORT_WIDTH_IN,
  height   = imp_plot_height(imp_moss_change) + imp_plot_height(imp_plant_change) + LEGEND_HEIGHT + COMBINED_PADDING - 0.3, # Slightly reduce height to avoid clipping the legend
  dpi      = 300, bg = "white"
)
```